plyranges: a grammar for transforming genomics data

### 2018-04-30 paper.Rmd
update figures

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by in the aforementioned Bioconductor packages and are not easily 
extended.

@@

The `plyranges` DSL is built on the core Bioconductor data structure
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example, a GRanges can represent a gene with
its chromosomal coordinates, exonic structure, and a count column
summarized from an RNA-seq experiment.

@@

\begin{table}[ht]
\label{tab:grammar}
\centering
\begin{tabular}{|l|l|p{4cm}|}
  \hline
Operation Type & Verb &  Description \\ 
  \hline
  Arithmetic &  `mutate()` & modifies any column \\
  Arithmetic & `select()` & select columns \\
  Arithmetic & `arrange()` & sort by columns \\
  Arithmetic & `set\_width()` & modifies width \\
  Arithmetic & `stretch()` & extend range by fixed amount \\
  Arithmetic & `shift\_(direction)` & shift coordinates \\
  Arithmetic & `flank\_(direction)` & generate flanking regions \\
  Arithmetic & `\%setop\%` & row-wise set operations \\
  Restriction & `filter()` & subset rows \\
  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
  Merging & `(setop)\_ranges` & range-wise set operations \\
  Aggregation & `summarise()` & aggregate over column(s) \\ 
  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
  Modifier & `group\_by()` & partition by column(s)  \\ 
  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
  Modifier & `anchor\_direction()` & fix coordinates \\
   \hline
\end{tabular}
\caption{Overview of `plyranges` grammar.}
\end{table}

@@

```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators. Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively). An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input. This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
knitr::include_graphics("olaps-figure.png")
```

@@

Here we provide examples on how to use the `plyranges` DSL to construct
genomic data workflows and highlight its interoperability with existing
Bionconductor packages.

@@

```{r max-score-region}
chr10_max_score_region <- chr10_scores %>%
  filter(score == max(score)) %>% 
  anchor_center() %>%
  set_width(5000)
```

>>>>

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

@@

The `plyranges` DSL is built on the core Bioconductor data structure
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment (figure \ref{fig:GRanges}).

```{r GRanges, echo = FALSE, out.width="400pt", fig.cap="An illustration of the GRanges data model for a sample from an RNA-seq experiment. The core components of the data model inlcude a seqnames column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded). Metadata are included as columns to the right of the dotted line as annotations (gene_id) or range level covariates (GC score and count)."}
knitr::include_graphics("diagrams/diagrams.001.png")
```

@@

*

\begin{table}[ht]
\label{tab:grammar}
\centering
\begin{tabular}{|l|l|p{4cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
  Arithmetic &  {\tt mutate()} & modifies any column \\
  Arithmetic & `select()` & select columns \\
  Arithmetic & `arrange()` & sort by columns \\
  Arithmetic & `set\_width()` & modifies width \\
  Arithmetic & `stretch()` & extend range by fixed amount \\
  Arithmetic & `shift\_(direction)` & shift coordinates \\
  Arithmetic & `flank\_(direction)` & generate flanking regions \\
  Arithmetic & `\%setop\%` & row-wise set operations \\
  Restriction & `filter()` & subset rows \\
  Restriction & `filter\_by\_overlaps()` & subset by overlap \\
  Restriction & `filter\_by\_non\_overlaps()` & subset by no overlap \\
  Merging & `join\_overlap\_()` & merge by overlapping ranges \\
  Merging & `join\_(follow/precedes/nearest)` & merge nearest neighbours \\
  Merging & `(setop)\_ranges` & range-wise set operations \\
  Aggregation & `summarise()` & aggregate over column(s) \\ 
  Aggregation & `disjoin\_ranges()` & aggregate column(s) over the union of end coordinates \\
  Aggregation & `reduce\_ranges()` & aggregate column(s) by merging overlapping and neighbouring ranges \\
  Modifier & `group\_by()` & partition by column(s)  \\ 
  Modifier & `group\_by\_overlaps()` & partition by overlaps \\
  Modifier & `anchor\_direction()` & fix coordinates \\
   \hline
\end{tabular}
\caption{Overview of `plyranges` grammar. Expand here (bold to emphasise
dplyr adaptations)}
\end{table}

@@

```{r olaps-fig, echo = FALSE, out.width="400pt", fig.cap="Illustration of the three overlap join operators. Each join takes query and subject range as input (black and grey rectangles, respectively) with associated metadata (pink and orange rectangles respectively). An index for the join is computed via returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query or (take the intersect) of the query ranges with subject ranges and return metadata associated with each input. This principle is gnerally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
knitr::include_graphics("diagrams/diagrams.002.png")
```

@@

(change this)
Here we provide examples on how to use the `plyranges` DSL to construct
genomic data workflows and highlight its interoperability with existing
Bionconductor packages.

@@

```{r max-score-region}
chr10_max_score_region <- chr10_scores %>%
  filter(score == max(score)) %>% 
  anchor_center() %>%
  mutate(width = 5000)
```
